---
title: "plot_gc"
author: "Aïsha Schuhegger, Helene Wetekam"
date: "2/15/2024"
output: html_document
---

# This script is for plotting the genetic correlations found by LDSC and HDL as pheatmaps. 
# Load libraries
```{r}
library(ggplot2)
library(reshape2)
```

# Read in the separate tsv files (output_ld_regression tsv files)
```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Input folder
folder_path_ldsc <- "../output_ld_regression"
folder_path_hdl <- "../output_hdl"

# Output folder
folder_path_plots <- "../plots/"
folder_heatmaps_ldsc <- "../plots/heatmaps_ldsc"
folder_heatmaps_hdl <- "../plots/heatmaps_hdl"

# Check if the folder exists
if (!dir.exists(folder_path_plots)) {
  # If the folder does not exist, create it
  dir.create(folder_path_plots, recursive = TRUE)
  print(paste("Folder", folder_path_plots, "created."))
} else {
  print(paste("Folder", folder_path_plots, "already exists."))
}

# Check if the folder exists
if (!dir.exists(folder_heatmaps_ldsc)) {
  # If the folder does not exist, create it
  dir.create(folder_heatmaps_ldsc, recursive = TRUE)
  print(paste("Folder", folder_heatmaps_ldsc, "created."))
} else {
  print(paste("Folder", folder_heatmaps_ldsc, "already exists."))
}

# Check if the folder exists
if (!dir.exists(folder_heatmaps_hdl)) {
  # If the folder does not exist, create it
  dir.create(folder_heatmaps_hdl, recursive = TRUE)
  print(paste("Folder", folder_heatmaps_hdl, "created."))
} else {
  print(paste("Folder", folder_heatmaps_hdl, "already exists."))
}

# Read LDSC files
matrix_ldsc <- read.table(paste(folder_path_ldsc, "/matrix_df_ldsc.tsv", sep=""), header=TRUE)
matrix_woCorrection_ldsc <- read.table(paste(folder_path_ldsc, "/matrix_df_woCorrection_ldsc.tsv", sep=""), header=TRUE)
matrix_corr_sign_ldsc <- read.table(paste(folder_path_ldsc, "/matrix_df_correction_sign_ldsc.tsv", sep=""), header=TRUE)
# filter out GWAS studies where diag=0 of matrix_corr_sign_ldsc 
diag_vals_sign <- diag(as.matrix(median_matrix_sign))
matrix_corr_sign_ldsc_diagfiltered <- matrix_corr_sign_ldsc[diag_vals_sign >= 0.995 & diag_vals_sign <= 1.005, diag_vals_sign >= 0.995 & diag_vals_sign <= 1.005]

# Read HDL files
matrix_hdl <- read.table(paste(folder_path_hdl, "/matrix_df_hdl.tsv", sep=""), header=TRUE)
matrix_woCorrection_hdl <- read.table(paste(folder_path_hdl, "/matrix_df_woCorrection_hdl.tsv", sep=""), header=TRUE)
matrix_corr_sign_hdl <- read.table(paste(folder_path_hdl, "/matrix_df_correction_sign_hdl.tsv", sep=""), header=TRUE)
```

#Pheatmap LDSC
```{r}
#install.packages("pheatmap")
library(pheatmap)
my_palette <- colorRampPalette(c("firebrick4", "white", "royalblue4"))(100)

# Create pheatmap with corrected correlation values (>1 values = 1, <-1 values = -1)

# Without hclust 
pheatmap_plot_withoutHclust <- pheatmap(matrix_ldsc,
                          cluster_rows = FALSE,  
                          cluster_cols = FALSE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - corrected gc values (LDSC)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 3,      # Set fontsize for row labels
                          fontsize_col = 3,      # Set fontsize for column labels
                          cellwidth = 3,        # Set cell width
                          cellheight = 3,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_ldsc, "/pheatmap_plot_withoutHclust_ldsc.png", sep=""), plot = pheatmap_plot_withoutHclust, width = 7, height = 7, units = "in", dpi = 300)

# With hclust 
pheatmap_plot_hclust <- pheatmap(matrix_ldsc,
                          cluster_rows = FALSE,  
                          cluster_cols = TRUE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - corrected gc values and hclust (LDSC)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 3,      # Set fontsize for row labels
                          fontsize_col = 3,      # Set fontsize for column labels
                          cellwidth = 3,        # Set cell width
                          cellheight = 3,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_ldsc, "/pheatmap_plot_hclust_ldsc.png", sep="") , plot = pheatmap_plot_hclust, width = 7, height = 7, units = "in", dpi = 300)

# Create pheatmap without corrected correlation values (>1 and <-1 values are not corrected)

# Without hclust 
pheatmap_plot_woCorrection_withoutHclust <- pheatmap(matrix_woCorrection_ldsc,
                          cluster_rows = FALSE,  
                          cluster_cols = FALSE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - uncorrected gc values (LDSC)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 3,      # Set fontsize for row labels
                          fontsize_col = 3,      # Set fontsize for column labels
                          cellwidth = 3,        # Set cell width
                          cellheight = 3,       # Set cell height
                          col = my_palette       # Color scheme
)
ggsave(paste(folder_heatmaps_ldsc, "/pheatmap_plot_woCorrection_withoutHclust_ldsc.png", sep=""), plot = pheatmap_plot_woCorrection_withoutHclust, width = 7, height = 7, units = "in", dpi = 300)

# With hclust 
pheatmap_plot_woCorrection_hclust <- pheatmap(matrix_woCorrection_ldsc,
                          cluster_rows = FALSE,  
                          cluster_cols = TRUE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - uncorrected gc values and hclust (LDSC)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 3,      # Set fontsize for row labels
                          fontsize_col = 3,      # Set fontsize for column labels
                          cellwidth = 3,        # Set cell width
                          cellheight = 3,       # Set cell height
                          col = my_palette       # Color scheme
)
ggsave(paste(folder_heatmaps_ldsc, "/pheatmap_plot_woCorrection_hclust_ldsc.png", sep=""), plot = pheatmap_plot_woCorrection_hclust, width = 7, height = 7, units = "in", dpi = 300)

# Create pheatmap for corrected gc and significant p-values

# Without hclust 
pheatmap_plot_corr_sign_withoutHclust <- pheatmap(matrix_corr_sign_ldsc_diagfiltered,
                          cluster_rows = FALSE,  
                          cluster_cols = FALSE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - corrected gc values\nand significant p-values (LDSC)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 5,      # Set fontsize for row labels
                          fontsize_col = 5,      # Set fontsize for column labels
                          cellwidth = 5,        # Set cell width
                          cellheight = 5,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_ldsc, "/pheatmap_plot_corr_sign_withoutHclust_ldsc.png", sep=""), plot = pheatmap_plot_corr_sign_withoutHclust, width = 7, height = 7, units = "in", dpi = 300)

# With hclust 
pheatmap_plot_corr_sign_hclust <- pheatmap(matrix_corr_sign_ldsc_diagfiltered,
                          cluster_rows = FALSE,  
                          cluster_cols = TRUE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - corrected gc values,\nsignificant p-values and hclust (LDSC)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 5,      # Set fontsize for row labels
                          fontsize_col = 5,      # Set fontsize for column labels
                          cellwidth = 5,        # Set cell width
                          cellheight = 5,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_ldsc, "/pheatmap_plot_corr_sign_hclust_ldsc.png", sep=""), plot = pheatmap_plot_corr_sign_hclust, width = 7, height = 7, units = "in", dpi = 300)
```

#Pheatmap HDL
```{r}
#install.packages("pheatmap")
library(pheatmap)
my_palette <- colorRampPalette(c("firebrick4", "white", "royalblue4"))(100)

# Create pheatmap with corrected correlation values (>1 values = 1, <-1 values = -1)

# Without hclust 
pheatmap_plot_withoutHclust <- pheatmap(matrix_hdl,
                          cluster_rows = FALSE,  
                          cluster_cols = FALSE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - corrected gc values (HDL)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 4,      # Set fontsize for row labels
                          fontsize_col = 4,      # Set fontsize for column labels
                          cellwidth = 4,        # Set cell width
                          cellheight = 4,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_hdl, "/pheatmap_plot_withoutHclust_hdl.png", sep=""), plot = pheatmap_plot_withoutHclust, width = 7, height = 7, units = "in", dpi = 300)

# With hclust 
pheatmap_plot_hclust <- pheatmap(matrix_hdl,
                          cluster_rows = FALSE,  
                          cluster_cols = TRUE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - corrected gc values and hclust (HDL)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 4,      # Set fontsize for row labels
                          fontsize_col = 4,      # Set fontsize for column labels
                          cellwidth = 4,        # Set cell width
                          cellheight = 4,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_hdl, "/pheatmap_plot_hclust_hdl.png", sep="") , plot = pheatmap_plot_hclust, width = 7, height = 7, units = "in", dpi = 300)

# Create pheatmap without corrected correlation values (>1 and <-1 values are not corrected)

# Without hclust 
pheatmap_plot_woCorrection_withoutHclust <- pheatmap(matrix_woCorrection_hdl,
                          cluster_rows = FALSE,  
                          cluster_cols = FALSE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - uncorrected gc values (HDL)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 4,      # Set fontsize for row labels
                          fontsize_col = 4,      # Set fontsize for column labels
                          cellwidth = 4,        # Set cell width
                          cellheight = 4,       # Set cell height
                          col = my_palette       # Color scheme
)
ggsave(paste(folder_heatmaps_hdl, "/pheatmap_plot_woCorrection_withoutHclust_hdl.png", sep=""), plot = pheatmap_plot_woCorrection_withoutHclust, width = 7, height = 7, units = "in", dpi = 300)

# With hclust 
pheatmap_plot_woCorrection_hclust <- pheatmap(matrix_woCorrection_hdl,
                          cluster_rows = FALSE,  
                          cluster_cols = TRUE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - uncorrected gc values and hclust (HDL)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 4,      # Set fontsize for row labels
                          fontsize_col = 4,      # Set fontsize for column labels
                          cellwidth = 4,        # Set cell width
                          cellheight = 4,       # Set cell height
                          col = my_palette       # Color scheme
)
ggsave(paste(folder_heatmaps_hdl, "/pheatmap_plot_woCorrection_hclust_hdl.png", sep=""), plot = pheatmap_plot_woCorrection_hclust, width = 7, height = 7, units = "in", dpi = 300)

# Create pheatmap for corrected gc and significant p-values

# Without hclust 
pheatmap_plot_woCorrection_withoutHclust <- pheatmap(matrix_corr_sign_hdl,
                          cluster_rows = FALSE,  
                          cluster_cols = FALSE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - corrected gc values\nand significant p-values (HDL)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 6,      # Set fontsize for row labels
                          fontsize_col = 6,      # Set fontsize for column labels
                          cellwidth = 7,        # Set cell width
                          cellheight = 7,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_hdl, "/pheatmap_plot_corr_sign_withoutHclust_hdl.png", sep=""), plot = pheatmap_plot_woCorrection_withoutHclust, width = 7, height = 7, units = "in", dpi = 300)

# With hclust 
pheatmap_plot_woCorrection_hclust <- pheatmap(matrix_corr_sign_hdl,
                          cluster_rows = FALSE,  
                          cluster_cols = TRUE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 7,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap - corrected gc values,\nsignificant p-values and hclust (HDL)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 6,      # Set fontsize for row labels
                          fontsize_col = 6,      # Set fontsize for column labels
                          cellwidth = 7,        # Set cell width
                          cellheight = 7,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_hdl, "/pheatmap_plot_corr_sign_hclust_hdl.png", sep=""), plot = pheatmap_plot_woCorrection_hclust, width = 7, height = 7, units = "in", dpi = 300)
```


Check diagonal values of the uncorrected gc values for LDSC
-> No need to do this for HDL, because we added gc(gwas1,gwas1) manually afterwards (HDL cannot be called on gwas1 with gwas1)
```{r, comment=NA}
# Check diagonal values of the uncorrected gc values -> they should be 1 as there is full correlation
matrix_woCorrection <- as.matrix(matrix_woCorrection_ldsc)
diagonal_values <- diag(as.matrix(matrix_woCorrection_ldsc))

# Initialize counters
correlation_diag_yes <- 0
correlation_diag_no <- 0

# Define tolerance levels
tolerance_lower <- 0.995
tolerance_upper <- 1.005

# Iterate over each value in the diagonal correlation values 
for (value in diagonal_values) {
  # Check if the value falls within the tolerance range
  if (value < tolerance_upper && value > tolerance_lower) {
    correlation_diag_yes <- correlation_diag_yes + 1
  } else {
    correlation_diag_no <- correlation_diag_no + 1 
  }
}

# Print the counts
cat(paste("Within tolerance range:", correlation_diag_yes, "\n"))
cat(paste("Outside tolerance range:", correlation_diag_no))
```

Determine a threshold in order to create a more dense pheatmap for LDSC
-> Check all columns of the matrix with adjustments and check how many values are 0
-> Choose median as threshold instead of mean as median is not influenced by outliers 
```{r}
sum_of_notzeros_sign <- colSums(matrix_corr_sign_ldsc != 0)
median_of_notzeros_sign <- median(sum_of_notzeros_sign)
median_matrix_sign <- matrix_corr_sign_ldsc[sum_of_notzeros_sign > median_of_notzeros_sign, sum_of_notzeros_sign > median_of_notzeros_sign]

# Only those where the diag value = 1or in tolerance range 
diag_vals_sign <- diag(as.matrix(median_matrix_sign))
filtered_median_matrix_sign <- median_matrix_sign[diag_vals_sign >= 0.995 & diag_vals_sign <= 1.005, diag_vals_sign >= 0.995 & diag_vals_sign <= 1.005]

# Without hclust 
pheatmap_median_sign <- pheatmap(filtered_median_matrix_sign,
                          cluster_rows = FALSE,  
                          cluster_cols = FALSE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 10,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap \n median as threshold (p<0.05) (LDSC)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 6.5,      # Set fontsize for row labels
                          fontsize_col = 6.5,      # Set fontsize for column labels
                          cellwidth = 10,        # Set cell width
                          cellheight = 10,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_ldsc, "/pheatmap_median_sign_ldsc.png", sep=""), plot = pheatmap_median_sign, width = 7.5, height = 7.5, units = "in", dpi = 300)

# With shclust 
pheatmap_median_hclust_sign <- pheatmap(filtered_median_matrix_sign,
                          cluster_rows = FALSE,  
                          cluster_cols = TRUE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 10,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap \n median as threshold and clustering (p<0.05) (LDSC)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 6.5,      # Set fontsize for row labels
                          fontsize_col = 6.5,      # Set fontsize for column labels
                          cellwidth = 10,        # Set cell width
                          cellheight = 10,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)

ggsave(paste(folder_heatmaps_ldsc, "/pheatmap_median_hclust_sign_ldsc.png", sep=""), plot = pheatmap_median_hclust_sign, width = 6, height = 7.5, units = "in", dpi = 300)
```

Determine a threshold in order to create a more dense heatmap for HDL
```{r}
sum_of_notzeros_sign <- colSums(matrix_corr_sign_hdl != 0)
median_of_notzeros_sign <- median(sum_of_notzeros_sign) 
median_matrix_sign <- matrix_corr_sign_hdl[sum_of_notzeros_sign > median_of_notzeros_sign, sum_of_notzeros_sign > median_of_notzeros_sign]

# Without hclust 
pheatmap_median_sign <- pheatmap(median_matrix_sign,
                          cluster_rows = FALSE,  
                          cluster_cols = FALSE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 10,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap \n median as threshold (p<0.05) (HDL)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 6.5,      # Set fontsize for row labels
                          fontsize_col = 6.5,      # Set fontsize for column labels
                          cellwidth = 10,        # Set cell width
                          cellheight = 10,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)
ggsave(paste(folder_heatmaps_hdl, "/pheatmap_median_sign_hdl.png", sep=""), plot = pheatmap_median_sign, width = 7.5, height = 7.5, units = "in", dpi = 300)

# With shclust 
pheatmap_median_hclust_sign <- pheatmap(median_matrix_sign,
                          cluster_rows = FALSE,  
                          cluster_cols = TRUE, 
                          annotation_col = NULL,  # Use column names as annotations at the top
                          border_col = NA,
                          annotation_row = NULL, 
                          fontsize = 10,          # Set fontsize for labels
                          main = "Genetic Correlation Heatmap \n median as threshold and clustering (p<0.05) (HDL)", 
                          angle_col = 90,       
                          angle_row = 0,        
                          fontsize_row = 6.5,      # Set fontsize for row labels
                          fontsize_col = 6.5,      # Set fontsize for column labels
                          cellwidth = 10,        # Set cell width
                          cellheight = 10,       # Set cell height
                          col = my_palette,       # Color scheme
                          breaks = seq(-1, 1, length.out = 101)  # Define breaks from -1 to 1
)

ggsave(paste(folder_heatmaps_hdl, "/pheatmap_median_hclust_sign_hdl.png", sep=""), plot = pheatmap_median_hclust_sign, width = 6, height = 7.5, units = "in", dpi = 300)
```



